Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-03-16 demystify.tex
Updates from Overleaf

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}
\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc} 
\usepackage{anyfontsize}
\usepackage{sectsty}

@@

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system, i.e., not by the CS department. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS (in some cases the committee can involve affiliated faculty from different disciplines). These faculty have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that many schools, including GMU, have separate adcoms for MS programs.

Each application is assigned to 2--3 faculty members, who will evaluate your profile and reach a consensus.  Note that while the assigned reviewers are the main ones deciding your application, other faculty in the department will likely have access to your application and can provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various faculty titles). This provides junior faculty the opportunities to recruit students. Some members can be tenured faculty. The chair of the committee will likely a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty.

@@

  Another sad reality is that some professors are unwilling to write letters or write weak ones because they do not want (good) students to go abroad or only go to places where they want the students to go to. If you are in this situation, you should find someone else to write for you.
  \tcblower

  Sometimes students would go through great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, the letters are generic and carry little value, in fact, \red{red flags}. Moreover, a top professor in Vietnam might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and just get letters from \emph{any} professors/supervisors who knows you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized
  letter about your own research ability from someone who is less
  well-known than a generic/weak letter from a well-known person.

@@

What can you do about this? You can explain about your school in your SOP (and also ask your LOR writer to do that). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Grades} Having good GPA is good. However if your school is not well-known, having top grades or rankings
usually will not help because we cannot evaluate them.

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).
So what to do with this? As mentioned in \autoref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

Note that while having good grades at unknown school might not help,
having bad grades will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses (e.g., Math and CS).

@@

\paragraph{GRE} Most good CS PhD programs in the US \textbf{no longer require GREs}, so you don't need to take them. However, they might be useful for international students from programs we are not familiar with.  If you have good GRE scores then you should include them in your application (but don't expect them to make much difference).  
\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test. Just do well enough to pass minimum requirement set by the university, which is required for TA (\autoref{sec:funding}). Just as with grades, having a high school in these tests might not help, but having too low scores can a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\paragraph{Work Experiences} Work experiences, especially if they are not research-based, do not carry much weight. So do not spend much time talking about them in your SOP (and similarly, LoR from your supervisor for non-research experience will not count much).

However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab (e.g., Microsoft, DoE, VinAI in Vietnam, etc) (\autoref{sec:research-ability}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

\section{CV/Resume}
The CV/Resume should provide a summary of the applicant's achievements.  It should allow the reviewers to quickly scan to identify major achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English. 
In general, just treat the interview as an informal chat, but be prepared to talk about your research experiences and interests (so do have \emph{elevator pitch} about your experiences and research). 
A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). Some prof. can ask you to solve a problem, e.g., a coding problem (so you might want to ask if you need to prepare for that).
During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

@@

Thus, if your goal is Ph.D. in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities. 
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. All schools listed in \autoref{sec:ranking} have PhD programs in CS, so you can start there.

@@

If you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you really need to put effort into it and do it the \emph{right way}.

@@

Below is a good example that I would reply to.

@@

>>>>

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1.5in]{geometry}
\usepackage{soul}
% \usepackage[small,compact]{titlesec} %very powerful
\usepackage[most]{tcolorbox}
% \setsecnumdepth{subsection}
% \setcounter{tocdepth}{3}
\usepackage{enumitem}
\usepackage{epigraph}
\usepackage{cite}
\usepackage{caption}
\captionsetup{font=small}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{wrapfig}
\setlength\intextsep{0pt} % remove extra space above and below in-line float
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  citecolor=black,
  filecolor=black,
  linkcolor=blue,
  urlcolor=blue,
}
\usepackage{booktabs}
%\usepackage[dvipsnames]{xcolor}
\usepackage{tikz}
\usetikzlibrary{calc} 
\usepackage{anyfontsize}
\usepackage{sectsty}

@@

After you submit your PhD application (usually around December), it will first be screened
for general requirements, e.g., did you submit your transcripts and standard scores? did your reference writers submit their letters? Usually, this screen process is done through a central university system, i.e., not by faculty from the CS department. If you pass this screen, your application will be forwarded to the CS department for further evaluation. If you don't,  the system will tell you what is missing and what you need to do. So pay attention to your email and check your application status regularly.

@@

Then, applications are reviewed by a \emph{PhD admission committee} (\textbf{adcom}) that consists of faculty members in CS\footnote{In some cases the committee can involve affiliated faculty from different disciplines.}. These adcom members have a wide-range of expertise and background to ensure diverse perspectives in the evaluation process. The size and the review load of the adcom depends on the department size. At GMU, the PhD adcom typically has 15--20 faculty, and each committee member is assigned to review 25--30 applications. Note that most large schools, including GMU, have separate adcoms for MS programs.

Each application is assigned to about 3 faculty members, who will evaluate your profile and reach a consensus.  While the assigned reviewers are the main ones deciding your application, other faculty in the department will likely have access to your application and can provide inputs and opinions on your profile.

The PhD adcom typically involves assistant professors in the department (see \autoref{sec:faculty-types} for various types of faculty). This provides junior faculty the opportunities to recruit students. Some members are tenured faculty. The chair of the committee will likely a senior faculty, but they will not review individual applications and instead assign them to committee members. The chair will look at various factors such as research interests or mentioning faculty names to assign the applications to appropriate faculty, e.g., I am often assigned to review applicants interested in software engineering.

@@

  Several international students mentioned that some professors are unwilling to write letters or write weak ones because they do not want (good) students to go abroad or only go to places where they want the students to go to. If you are in this situation, you should find someone else to write for you.
  \tcblower

  Sometimes students would go through great length just to get letters from well-known professors in their school (e.g., department head). But as mentioned, if these professors do not know you, their letters would likely be generic and carry little value, in fact, \red{red flags}. Moreover, a top professor at your university might not be well-known to US faculty (see more details in \autoref{sec:your-school}). So save the trouble and  get letters from \emph{any} professors/supervisors who knows you well and can write a good letter about \emph{your} research ability. It's better to have a good personalized
  letter about your own research ability from someone who is less
  well-known than a generic/weak letter from a well-known person.

@@

\textbf{Advice}: You can explain about your school in your SOP (and also ask your LOR writer to do that). Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Grades} Having good GPA is good. However if your school is not well-known, having top grades or rankings
usually will not help because we cannot evaluate them.  This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the adcom).

\textbf{Advice}: As with school reputation, you can put a note about this in your statement and ask your LoR writers to talk about it.
Note that while having good grades at unknown school might not help,
having bad grades (e.g., < 3.0) will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses, e.g., Math and CS.

@@

\paragraph{GRE} Most good CS PhD programs in the US (e.g., R1 universities) \textbf{no longer require GREs}, so you don't need to take them.

\textbf{Advice}: In general, don't waste your time taking GRE.  However, if you took it and have good scores then you can include and talk about them in your application (but don't expect them to make much difference).

\paragraph{English Test} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-GMU#standard-tests-waiver-eligible-countries}{these}, you will need to
take standardized English test.

\textbf{Advice}: Just do well enough to pass minimum requirement set by the university, which is required for TA (\autoref{sec:funding}). Just as with grades (\autoref{sec:your-school}) and GRE, having a high scores in these tests might not help, but having too low scores can a \red{red flag} and sometimes results in an automatic rejection (\autoref{sec:why-rejected}), e.g., below the minimum requirement.

@@

\paragraph{Work Experiences} Work experiences, especially if they are not research-based, do not carry much weight. Similarly, LoR from your supervisor for non-research experience will not count much.

\textbf{Advice}: do not spend much time talking about non-research related job in your SOP.  
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab (e.g., Microsoft, DoE, VinAI in Vietnam, etc) (\autoref{sec:research-ability}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potentials in their LOR (\autoref{sec:lor}).

\section{CV/Resume}
The CV/Resume provides a summary of the applicant's achievements.

\textbf{Advice}: Prepare your CV in such away that it allows the reviewers to quickly scan to identify major achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

@@

Typically, an interview takes about 15--30 minutes, and one important aspect of evaluation is your ability to effectively communicate, including speaking and understanding English.

\textbf{Advice}: In general, just treat the interview as an informal chat, but be prepared to talk about your research experiences and interests (so do have \emph{elevator pitch} about your experiences and research). 
A professor might want to know more about your research background or even ask you to read a paper (so be sure to do it!). Some prof. can ask you to solve a problem, e.g., a coding problem, so you might want to ask if you need to prepare for that.
During the interview, you will have the opportunity to ask questions about the university. It's essential to prepare some thoughtful questions, similar to a job interview.

@@

Thus, if your goal is Ph.D. in CS, you have to aim only for schools offering such a degree.  %This might sound obvious but many students (even those in the US) get confused due to the large number of universities. 
While this can be confusing due to the large number of universities in the US, a little research, e.g., searching for PhD in CS from the school website, will help you find out. Schools listed in \autoref{sec:ranking} have PhD programs in CS, so you can start there.

@@

On the other hand, if you want to contact a prof. to ask about \emph{research opportunities}, or \emph{GTA/GRA} support, then \emph{yes}, I believe you should---it is \emph{worth it}. However, you really need to put effort into it and do it the \emph{right way}.

@@

Below is a good example that I would likely reply.

@@

% \section{How much time did you spend looking at an application?}\label{sec:time-spent}
% Short answer: less than 5 mins for quick rejection; 15+ mins for an acceptance.  On average, about 10 mins.

% The GMU evaluation system allows us to download a single pdf that consists of a summary page (e.g., school info, GPA, testscores), then transcripts, standard test results, LoRs, CV, SOP, writing samples So I typically read the application in this order.  Then I make notes on what I've seen, basically evidence that support my evaluation.

% I first look at the summary for school, GPA, est scores (essentially looking for very low GPA or testscores that do not meet the minimum requirements. Then I skim on transcripts look for low/failing grades on CS and related courses.  For these I make comments such as "unknown international school with good GPA, too many low grades on CS and related courses" or "good GPA from top school XYZ". I do not look at details on tests (the main scores are already given in the first summary page).

% Then I look at LoRs.  I read carefully good LoRs and quickly skim over bad or generic ones (very easy to tell, you don't even need much experiences to distinguish between a strong LoRs from the rest). Comment examples For LoRs include "very strong LoRs from well-known prof. XYZ talking about research experiences" or "generic LoRs, mostly about course work".  I also note any unusual things in the LoRs.

% Then the SOP, which I read carefully (of course if it is bad then it would a quick read). Then my notes for SOP include things such as "SOP mostly on course work, no clear motivation, generic" , or "very enjoyable SOP talking about XYZ, customized for GMU, mentioned profs. X,Y,Z".  If the SOP mentioned about publications then I'll make notes such as "has X papers at top venues" or "has X papers but at unknown places".  I will not go online search for the papers (unless they upload them as sample papers) .

% I also skim over sampling papers, essentially looking over the abstract to see if reads well.  I would make notes such as "have a paper under submission to unknown venue, but reads well".

%At GMU we use an evaluation system that allows us to download a single, large pdf that consists of everything submitted by the applicant, e.g., GPA, transcripts, test scores, CV, LoRs, SOP, sample writings, etc.   The first couple of pages show the summary including school, major, GPA, test scores.
